---
title: "5. Agreement and ICC for Wide Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Agreement and ICC for Wide Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Scope

Agreement and reliability are related to correlation, but they are not the
same problem. Correlation describes co-movement. Agreement describes similarity
on the measurement scale itself. Reliability describes the proportion of
variation attributable to stable differences among subjects rather than to
measurement error or method disagreement.

This vignette focuses on `ccc()` for wide data and uses related agreement and
reliability functions as context:

- `ccc()`
- `ba()`
- `icc()`
- `ccc_rm_reml()` and `ccc_rm_ustat()`
- `icc_rm_reml()`
- `cia()` and `cia_rm()`

## Pairwise concordance and Bland-Altman analysis

Lin's concordance correlation coefficient combines precision and accuracy in a
single number. In `matrixCorr`, `ccc()` computes Lin's pairwise CCC for numeric
wide data and optionally returns large-sample confidence intervals. No formal
hypothesis test is implemented; inference is based on the estimate and its
confidence interval.

Bland-Altman analysis separates the agreement question into estimated bias and
limits of agreement.

```{r}
library(matrixCorr)

set.seed(40)
ref <- rnorm(50, mean = 100, sd = 10)
m1 <- ref + rnorm(50, sd = 2)
m2 <- ref + 1.2 + rnorm(50, sd = 3)

fit_ba <- ba(m1, m2)
fit_ccc <- ccc(data.frame(m1 = m1, m2 = m2), ci = TRUE)

print(fit_ba)
summary(fit_ccc)
```

The two summaries are complementary rather than redundant. `ccc()` gives a
single concordance coefficient, while `ba()` makes the scale of disagreement
explicit.

If you have at least 3 methods in wide form, `ba()` can now compute every
unordered Bland-Altman contrast directly:

```{r}
fit_ba_pairwise <- ba(data.frame(m1 = m1, m2 = m2, m3 = ref - 0.8 + rnorm(50, sd = 2.5)))
print(fit_ba_pairwise)
summary(fit_ba_pairwise)
```

## Pairwise ICC

`icc()` extends the wide-data reliability workflow in two directions. It can
return a pairwise matrix across method pairs, or it can return the overall
classical ICC table for the full set of methods.

```{r}
wide_methods <- data.frame(
  J1 = ref + rnorm(50, sd = 1.5),
  J2 = ref + 4.0 + rnorm(50, sd = 1.8),
  J3 = ref - 3.0 + rnorm(50, sd = 2.0),
  J4 = ref + rnorm(50, sd = 1.6)
)

fit_icc_pair <- icc(
  wide_methods,
  model = "twoway_random",
  type = "agreement",
  unit = "single",
  scope = "pairwise"
)

fit_icc_overall <- icc(
  wide_methods,
  model = "twoway_random",
  type = "agreement",
  unit = "single",
  scope = "overall",
  ci = TRUE
)

print(fit_icc_pair, digits = 2)
summary(fit_icc_pair)
print(fit_icc_overall)
```

## Pairwise versus overall ICC

This is the most important distinction in the ICC interface.

`scope = "pairwise"` answers:
"How reliable is each specific pair of methods?"

`scope = "overall"` answers:
"How reliable is the full set of methods when analysed jointly?"

Those are different quantities. The overall ICC cannot, in general, be
recovered by averaging the pairwise matrix.

## Consistency versus agreement

This simulation also includes systematic method bias, so it is a natural place
to contrast `type = "consistency"` with `type = "agreement"`.

```{r}
fit_icc_cons <- icc(
  wide_methods,
  model = "twoway_random",
  type = "consistency",
  unit = "single",
  scope = "overall",
  ci = FALSE
)

fit_icc_agr <- icc(
  wide_methods,
  model = "twoway_random",
  type = "agreement",
  unit = "single",
  scope = "overall",
  ci = FALSE
)

data.frame(
  type = c("consistency", "agreement"),
  selected_coefficient = c(
    attr(fit_icc_cons, "selected_coefficient"),
    attr(fit_icc_agr, "selected_coefficient")
  ),
  estimate = c(
    attr(fit_icc_cons, "selected_row")$estimate,
    attr(fit_icc_agr, "selected_row")$estimate
  )
)
```

Consistency discounts additive method shifts, whereas agreement penalises them.
When methods differ mainly by a systematic offset, consistency can therefore
look substantially better than agreement.

## Model, type, and unit

The classical ICC family is controlled by three arguments.

- `model` selects the one-way, two-way random, or two-way mixed formulation.
- `type` selects consistency or agreement.
- `unit` selects single-measure or average-measure reliability.

For pairwise ICC, average-measure output uses `k = 2` because each estimate is
based on exactly two methods. For overall ICC, average-measure output uses the
full number of analysed columns.

## Agreement indices implemented in matrixCorr

CCC, ICC and CIA address related but distinct agreement questions. CCC
quantifies agreement between measurements by combining precision and accuracy
(Lin, 1989). ICC expresses reliability or agreement through variance
components; its interpretation depends on the study design, model form, and
whether agreement or consistency is targeted (Shrout and Fleiss, 1979; McGraw
and Wong, 1996). CIA targets individual agreement or interchangeability by
comparing disagreement between methods with disagreement within methods
(Barnhart, Kosinski, and Haber, 2007; Barnhart et al., 2007).

These indices should not be treated as interchangeable without considering the
scientific question, data structure, and implemented estimator.

### CCC functions

`ccc()` is the simple wide-data CCC implementation. It estimates pairwise Lin
CCC values and, when `ci = TRUE`, returns Lin delta-method/Fisher-z confidence
intervals. It does not report a p-value or test decision.

For repeated-measures data, the package provides two CCC routes. `ccc_rm_reml()`
fits pairwise mixed models by REML and converts the fitted variance components
and fixed-effect bias term into repeated-measures CCC estimates. `ccc_rm_ustat()`
computes a nonparametric U-statistic repeated-measures CCC with optional
Fisher-z confidence intervals. These repeated-measures functions estimate
agreement parameters and optional confidence intervals; they do not implement a
formal hypothesis test of the CCC parameter. `ccc_rm_reml()` can use
boundary-aware likelihood-ratio tests for variance-component selection when
`vc_select = "auto"`, but those tests are model-selection diagnostics rather
than tests of CCC agreement.

### ICC functions

`icc()` computes classical ANOVA ICC forms for wide data. With
`scope = "pairwise"` it returns a pairwise matrix; with `scope = "overall"` it
returns the standard six-form overall coefficient table. The selected ICC form
is controlled by `model`, `type`, and `unit`, so the same numeric value should
not be interpreted without those design choices.

Pairwise `icc()` and repeated-measures `icc_rm_reml()` provide estimates and
optional confidence intervals without a formal test of a target ICC parameter.
For `icc(scope = "overall")`, the package reports ANOVA F statistics, degrees
of freedom, and p-values in the overall coefficient table. The implementation
and tests verify those reported p-values, but the package documentation does
not define a generic ICC agreement hypothesis for users; this vignette therefore
does not describe the overall ICC p-values as a generic test of agreement.

### CIA functions

`cia()` and `cia_rm()` are included here only as conceptual comparisons. CIA
targets individual agreement or interchangeability rather than the same
population concordance question targeted by CCC. In `cia()`, replicated readings
within each subject-method cell are required because the estimator compares
between-method disagreement with within-method replicate disagreement. The
function supports pairwise and overall CIA, optional reference-method scaling,
and optional confidence intervals by the selected inference method. No formal
hypothesis test is implemented; inference is based on the estimate and its
confidence interval.

`cia_rm()` is different from `cia()` because it targets matched repeated measurements
under conditions such as visits, raters, laboratories, treatments, or time
points, and it is not a technical-replicate estimator. It reports
condition-specific CIA estimates, optional confidence intervals, and a
homogeneity test for agreement across conditions. As implemented, the reported
test statistic is
`MS_method_time / MS_error`, with an upper-tail F-test p-value using
`df_method_time` and `df_error`. The null hypothesis documented by the function
is homogeneous agreement across conditions; the alternative is that agreement
changes across conditions.

### Function summary

In this package, the functions differ not only in the target index but also in
the supported data structure and inferential procedure. Some functions provide
estimates and confidence intervals only. A formal hypothesis test should only
be interpreted when the function explicitly implements and reports one.

```{r, echo = FALSE}
agreement_summary <- data.frame(
  Function = c(
    "`ccc()`",
    "`ccc_rm_reml()`",
    "`ccc_rm_ustat()`",
    "`icc()`",
    "`icc_rm_reml()`",
    "`cia()`",
    "`cia_rm()`"
  ),
  `Index family` = c(
    "Lin CCC",
    "Repeated-measures CCC",
    "Repeated-measures CCC",
    "Classical ICC",
    "Repeated-measures ICC",
    "CIA",
    "Repeated-measures CIA"
  ),
  `Target question` = c(
    "Pairwise concordance combining precision and accuracy",
    "Pairwise repeated-measures concordance from fitted variance components",
    "Pairwise repeated-measures concordance from U-statistic distances",
    "Reliability or agreement under the selected ICC model, type, and unit",
    "Pairwise repeated-measures reliability/agreement from fitted variance components",
    "Individual agreement/interchangeability relative to within-method replicate disagreement",
    "Individual agreement/interchangeability across matched repeated conditions"
  ),
  `Data/design supported` = c(
    "Numeric wide data; rows are paired observational units",
    "Long repeated-measures data; subject plus optional method/time structure",
    "Long repeated-measures data; balanced method/time coverage per pair",
    "Numeric wide data; pairwise or overall all-column scope",
    "Long repeated-measures data; subject plus optional method/time structure",
    "Long replicated method-comparison data with replicate identifiers",
    "Long matched repeated-measures data with one observation per subject-method-condition cell"
  ),
  `Estimation approach` = c(
    "Moment CCC with Lin delta-method/Fisher-z CI when requested",
    "REML mixed-model variance components and fixed-effect bias term",
    "Nonparametric U-statistic estimator with Fisher-z CI when requested",
    "Classical ANOVA mean-square formulas",
    "REML mixed-model variance components and fixed-effect bias term",
    "Method-of-moments disagreement ratios; optional bounded variance-component variant",
    "Categorical-condition ANOVA estimator"
  ),
  `Inference reported` = c(
    "Estimate; optional confidence interval",
    "Estimate; optional confidence interval; variance-component diagnostics",
    "Estimate; optional confidence interval",
    "Pairwise: estimate and optional CI. Overall: coefficient table with F statistic, df, p-value, and optional CI",
    "Estimate; optional confidence interval; variance-component diagnostics",
    "Estimate; optional confidence interval",
    "Estimate; optional confidence interval; homogeneity F statistic and p-value"
  ),
  `Formal hypothesis test implemented? Yes/No/Requires verification` = c(
    "No",
    "No for the CCC parameter; variance-component selection tests are not CCC tests",
    "No",
    "Requires verification for overall p-values; no pairwise ICC test is documented",
    "No for the ICC parameter; variance-component selection tests are not ICC tests",
    "No",
    "Yes: homogeneity of agreement across conditions"
  ),
  check.names = FALSE
)

knitr::kable(agreement_summary, escape = FALSE)
```


## Choosing among CCC, BA, ICC, and CIA

In practice these methods answer different questions.

- Use `ccc()` when one concordance coefficient per pair is the main target.
- Use `ccc_rm_ustat()` or `ccc_rm_reml()` when the concordance target is
  repeated-measures CCC rather than simple paired wide-data CCC.
- Use `ba()` when the size and direction of disagreement should be visible on
  the original measurement scale.
  Use `ba_rm()` for the repeated-measures Bland-Altman workflow.
- Use `icc()` when the target is reliability under a classical variance
  components interpretation.
- Use `icc_rm_reml()` when the target is repeated-measures ICC from the
  package's REML variance-component backend.
- Use `cia()` when replicated readings within subject-method cells are
  available and the target is individual agreement or interchangeability
  relative to within-method replicate disagreement.
- Use `cia_rm()` when the target is individual agreement across matched
  repeated conditions rather than technical replicates.

There is overlap in interpretation, but these are not interchangeable
estimators.

## References

Barnhart HX, Haber M, Lokhnygina Y, Kosinski AS. (2007). Comparison of
concordance correlation coefficient and coefficient of individual agreement in
assessing agreement. *Journal of Biopharmaceutical Statistics*, 17(4), 721-738.

Barnhart HX, Kosinski AS, Haber M. (2007). Assessing individual agreement.
*Journal of Biopharmaceutical Statistics*, 17(4), 697-719.

Carrasco JL, Jover L. (2003). Estimating the concordance correlation
coefficient: a new approach. *Computational Statistics & Data Analysis*, 47(4),
519-539.

Carrasco JL, Phillips BR, Puig-Martinez J, King TS, Chinchilli VM. (2013).
Estimation of the concordance correlation coefficient for repeated measures
using SAS and R. *Computer Methods and Programs in Biomedicine*, 109(3),
293-304.

Haber M, Gao J, Barnhart HX. (2010). Evaluation of agreement between
measurement methods from data with matched repeated measurements via the
coefficient of individual agreement. *Journal of Data Science*, 8, 457-469.

Lin L. (1989). A concordance correlation coefficient to evaluate
reproducibility. *Biometrics*, 45, 255-268.

McGraw KO, Wong SP. (1996). Forming inferences about some intraclass
correlation coefficients. *Psychological Methods*, 1(1), 30-46.

Pan Y, Gao J, Haber M, Barnhart HX. (2010). Estimation of coefficients of
individual agreement (CIA's) for quantitative and binary data using SAS and R.
*Computer Methods and Programs in Biomedicine*.

Shrout PE, Fleiss JL. (1979). Intraclass correlations: uses in assessing rater
reliability. *Psychological Bulletin*, 86(2), 420-428.
